---
title: "Reproducible code for: The rise of multidrug resistance in bacterial pathogens"
author: "Johnson T.F."
date: "2025-13-01"
output:
  pdf_document: default
  word_document: default
geometry: left=1cm,right=1cm,top=2cm,bottom=2cm
classoption: portrait
---

**Workflow to estimate rates of AMR (anti-microbial resistance) and MDR (multi-drug resistance)**


```{r setup, include=FALSE, warning=F, message = F, echo = F}
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 60), tidy = TRUE)
```

First we load necessary R packages: [ape 5.7.1](https://cran.r-project.org/web/packages/ape/index.html), [phytools 1.5.1](https://cran.r-project.org/web/packages/phytools/index.html), [geosphere 1.5-18](https://cran.r-project.org/web/packages/geosphere/index.html), [dplyr 1.1.2](https://dplyr.tidyverse.org/), [tidyr 1.3.0](https://tidyr.tidyverse.org/), [ggplot2 3.4.4](https://ggplot2.tidyverse.org/), [sf 1.0-13](https://r-spatial.github.io/sf/), [rnaturalearth 0.3.4](https://cran.r-project.org/web/packages/rnaturalearth/vignettes/rnaturalearth.html), [biscale 1.0.0](https://cran.r-project.org/web/packages/biscale/index.html), [cowplot 1.1.1](https://cran.r-project.org/web/packages/cowplot/vignettes/introduction.html), [ggtree 3.6.2](https://bioconductor.org/packages/devel/bioc/vignettes/ggtree/inst/doc/ggtree.html), [brms 2.20.4](https://paul-buerkner.github.io/brms/#:~:text=The%20brms%20package%20provides%20an,interface%20for%20performing%20regression%20analyses.), [stringr 1.5.1](https://stringr.tidyverse.org/), [ggpubr 0.6.0](https://rpkgs.datanovia.com/ggpubr/), [countrycode 1.5.0](https://cran.r-project.org/web/packages/countrycode/index.html), [raster 3.6-23](https://cran.r-project.org/web/packages/raster/index.html), [sp 2.1-1](https://cran.r-project.org/web/packages/sp/index.html)


After loading the R packages, we load two datasets (the records of resistance and a phylogeny at species level), and one function. We also introduce colour palettes for our figures. Generally, we use an increasing red palette to describe the relative multi-drug resistance. An increasing blue palette to depcit the change in resistance over time. 

```{r load}

library(phytools)
library(ape)
library(rnaturalearth)
library(rnaturalearthdata)
library(tidyverse)
library(countrycode)
library(geosphere)
library(rstan)
library(dplyr)
library(tidyr)
library(ggplot2)
library(sf)
library(biscale)
library(cowplot)
library(ggtree)
library(stringr)
library(raster)
library(sp)
library(ggdist)
library(ggpubr)
library(brms)
library(marginaleffects)
library(arrow)

df_save = read_parquet("DATA_AMR_jul2023_ISO3_LAT_LON 2.parquet")
tr_save = read.newick("species_level_tree (2).newick")

norm_range <- function(x){
  (x-min(x, na.rm = T))/(max(x, na.rm = T)-min(x, na.rm = T))
}

palette_hot = c("#f5f5f5", "#e6b6b6", "#d87676", "#c31f1f")
palette_cold = c( "#f5f5f5", "#c6c6c6","#999999","#6e6e6e")
palette_bi = c(
  "1-1" = "#f5f5f5", # low x, low y
  "2-1" = "#e6b6b6",
  "3-1" = "#d87676", 
  "4-1" = "#c31f1f", # high x, low y
  "1-2" = "#c6c6c6",
  "2-2" = "#ba9393",
  "3-2" = "#ae5f5f",
  "4-2" = "#9e1919",
  "1-3" = "#999999",
  "2-3" = "#907272", 
  "3-3" = "#874949",
  "4-3" = "#7a1414", 
  "1-4" = "#6e6e6e",# low x, high y
  "2-4" = "#675252",
  "3-4" = "#613535",
  "4-4" = "#570e0e"# high x, high y
)


```

After loading the data, we impose a series of restrictions, removing any rows missing: coordinates, year, species name, and information on whether resistance was detected across the 13 categories. We also only keep species/countries with atleast 10 observations (rows). 

```{r clean, warning=F, message = F}
df =  df_save
df = subset(df, !is.na(ISO.3)) #Remove all genomes with missing antibiotic information
df = subset(df, !is.na(centroid_lon)) #Remove all genomes with missing coordinates
df = subset(df, !is.na(centroid_lat))
print(nrow(df))

df = subset(df, !is.na(year)) #Remove all genomes with missing temporal information
print(nrow(df))

df = subset(df, !is.na(species)) #Remove all genomes with missing antibiotic information
print(nrow(df))

df = df[complete.cases(df[,c(9:21)]),] #Only keep complete cases
print(nrow(df))

geog_df = df %>%
  group_by(ISO.3) %>%
  summarise(N = n())
head(geog_df[order(-geog_df$N),]) #Countries with the most observations

world = ne_countries(scale = "medium", returnclass = "sf")
geog_df = left_join(world, geog_df, by = c("iso_a3" = "ISO.3"))

ggplot(data = geog_df) + #Plot observations over space
  geom_sf(aes(fill = N)) +
  scale_fill_gradient(name = "N", trans = "log10", breaks = c(1, 10, 100, 1000, 10000)) +
  theme_classic()

ggplot(data = geog_df) +  #Samples per capita over space
  geom_sf(aes(fill = N/(pop_est/1000000))) +
  scale_fill_gradient(name = "N/capita", trans = "log10", 
                      breaks = c(0.1, 1, 10, 100, 1000), labels = c("0.1", "1", "10", "100", "1000")) +
  theme_classic()

geog_time_df = df %>%
  group_by(ISO.3, year) %>%
  summarise(N = n())
ggplot(geog_time_df) + #Number of samples over time
     geom_col(aes(x = year, y = N), colour = "grey80", fill = "grey80") +
     labs(x = "", y = "N") +
     theme_classic()
geog_time_counts = data.frame(sort(table(geog_time_df$ISO.3)))
length(geog_time_counts[which(geog_time_counts$Freq == 1),1]) #Countries with observations in only one year
ggplot(geog_time_counts) + #Years of data per country
  geom_density(aes(x = Freq)) +
  labs(y = "Density (N)", x = "Years of data") +
  theme_classic()

geog_time_counts = left_join(world, geog_time_counts, by = c("iso_a3" = "Var1")) #Years of data over sapce
ggplot(data = geog_time_counts) +
  geom_sf(aes(fill = Freq)) +
  scale_fill_continuous(name = "Years of data") +
  theme_classic()


df = df[df$ISO.3 %in% geog_time_counts[which(geog_time_counts$Freq > 9),]$iso_a3, ] #Only keep data where there are atleast two observations per country

#Load environmental data
anti = read.csv("data_anti.csv")
anti$iso3 = countrycode(anti$country, "country.name", "iso3c")
anti$country = NULL

country_year_sum = df %>%
  group_by(year, ISO.3) %>%
  summarise(N = n())

country_year_sum = left_join(country_year_sum, anti, by = c("ISO.3" = "iso3", "year" = "year"))

ggplot(data = country_year_sum[which( #Plot number of samples by antibiotic use
  country_year_sum$year == 2018),]) +
  geom_point(aes(x = anti, y = N)) +
  scale_y_log10() +
  labs(x = "Antibiotic-use/capita", y = "N") +
  theme_classic()

ggplot(data = country_year_sum[which( #Plot number of samples by antibiotic use over different time periods
  country_year_sum$year == 2018 |
    country_year_sum$year == 2014 |
    country_year_sum$year == 2010 |
    country_year_sum$year == 2006),]) +
  geom_point(aes(x = anti, y = N)) +
  scale_y_log10() +
  labs(x = "Antibiotic-use/capita", y = "N") +
  theme_classic() +
  facet_wrap(year~.)

species_sum = df %>%
  group_by(species) %>%
  summarise(N = n())

ggplot(species_sum) + #Number of samples per species
  geom_col(aes(x = N, y = species)) +
  scale_x_log10() +
  labs(y = "", x = "N") +
  theme_classic()

species_year_sum = df %>%
  group_by(species, year) %>%
  summarise(N = n())
species_year_sum = species_year_sum %>%
  group_by(species) %>%
  summarise(N = n())

ggplot(species_year_sum) + #Number of years of data per species
  geom_col(aes(x = N, y = species)) +
  labs(y = "", x = "N") +
  theme_classic() 

df = df[df$species %in% species_year_sum[which(species_year_sum$N > 9),]$species, ] #

max(df$year) # 2023
df = subset(df, year > 1989)
df$year = df$year - max(df$year) #Rescale the values to improve model run times
colnames(df)[10] = "Betalactams"


```

Prior to modelling, we first must manipulate coordinates into a spatial covariance martix which specifies that neighbouring countires should have more similar rates of AMR. We use the Harvesine distances between country coordinate centroids. The Harvesine distnace acknowldges the curvature of the earth and calculates the number of meters between each country centorid. 

We also account for phylogenetic structure, under the expectation that AMR will be more similar in closely related species. Our phylogeny required manipulation to force it into an ultrametric structure.

For the modelling, we specify a multivariate mixed effect logistic regression, with 13 binary response variables, indicating signal of resistance (1) or not (0) to the following anti-microbials: Aminoglycosides, Betalactams, Polymyxins, Fosfomycin, Glycopeptides, Macrolides, Oxazolidinones, Phenicols, Quinolones, Rifampicin, Sulphonamides, Tetracyclines, Trimethoprim. We charecterise this resistance as a function of time (year fixed-effect). The data exhibits a hierarchical strucutre, which we capture by specifying correlated random intercepts and slopes for each baterial species and country. We also include two correlated effects: 1) specifying AMR in countries should covary according to pairwise distance from coordinate centroids; and 2) specifying AMR in bacteria should covary according to phylogenetic structure.


```{r stan_data, warning=F, message = F}

tr = tr_save
df$species = gsub(" ", "_", df$species)
df = df[df$species %in% tr$tip.label, ]


df$space_id = df$ISO.3
df$phylo_id = df$species
df$uni_id = 1:nrow(df)

# small sample 
subsampledf = df 
subsampledf = as.data.frame(subsampledf)
subsampledf = subsampledf %>% na.omit


GenerateMiniCov = function(df){
  #Produce a spatial covariance matrix
  spa_df = unique(df[,c("ISO.3", "centroid_lat", "centroid_lon")])
  spa_mat_trim = as.matrix(distm(spa_df[,c(3,2)], fun = distHaversine))/1000000
  # probably need kernel here 
  spa_mat_trim = norm_range(spa_mat_trim)
  spa_mat_trim = abs(spa_mat_trim - 1)
  spa_id = spa_df$ISO.3
  colnames(spa_mat_trim) = spa_id
  rownames(spa_mat_trim) = spa_id
  
  # phylo
  df = df[df$species %in% tr$tip.label, ]
  tr_trim = keep.tip(tr, unique(df$species))
  tr_trim_extend = phangorn::nnls.tree(cophenetic(tr_trim), tr_trim, rooted = TRUE) 
  tr_trim_extend$edge.length = tr_trim_extend$edge.length + 0.001
  phy_mat_trim = ape::vcv.phylo(tr_trim_extend, corr = T)
  
  # df
  df$space_id = df$ISO.3
  df$phylo_id = df$species
  df$uni_id = 1:nrow(df)
  
  outlist=list(space=spa_mat_trim,phylo=phy_mat_trim,df=df)
  return(outlist)
}


standata = GenerateMiniCov(subsampledf)

```


```{r stan_model, warning=F, message = F}

#form =
#  bf(mvbind(
#    Aminoglycosides, 
#    Betalactams, 
#    Polymyxins, 
#    Fosfomycin, 
#    Glycopeptides, 
#    Macrolides, 
#    Oxazolidinones, 
#    Phenicols, 
#    Quinolones, 
#    Rifampicin, 
#    Sulphonamides,
#    Tetracyclines,
#    Trimethoprim) ~ 
#      year + 
#      (1 + year|ISO.3) +
#      (1 + year|species) +
#      (1|gr(space_id, cov = space)) +
#      (1|gr(phylo_id, cov = phylo)) 
#  )  + set_rescor(FALSE)


#fit = brm(form, 
#           data = standata$df, 
#           data2 = list(space = standata$space, phylo = standata$phylo), 
#           chains = 4, cores = 4, iter = 500, warmup = 250, thin = 1, refresh = 5,
#           family = bernoulli)

#saveRDS(fit, "non_multivar.rds")
fit = readRDS("non_multivar.rds")
summary(fit)
```

### Results

```{r pp_check, fig.height=5, fig.width = 5, warning=F, message = F}

ab_list = c(
  "Aminoglycosides", 
  "Betalactams", 
  "Polymyxins", 
  "Fosfomycin", 
  "Glycopeptides", 
  "Macrolides", 
  "Oxazolidinones", 
  "Phenicols", 
  "Quinolones", 
  "Rifampicin", 
  "Sulphonamides",
  "Tetracyclines",
  "Trimethoprim")

pp_list = list()
for(a in ab_list){
  pp_list[[a]] = pp_check(fit, resp = a) + 
  theme_classic() +
  labs(title = a)
}

draw_n = 1000
tmp_df = data.frame(year = c(min(subsampledf$year): (0)))
draw = posterior_linpred(fit, tmp_df, ndraws = draw_n, re.form = NA, transform = T)
draw = matrix(draw, draw_n*34, 13)
meta = expand.grid(draws = c(1:draw_n), year = c(1990:2023))
draw = cbind(meta, draw)
colnames(draw)[3:ncol(draw)] = ab_list
draw_long = gather(draw, ab, prob, Aminoglycosides:Trimethoprim)


ab_probs = draw_long %>%
  group_by(ab, year) %>%
  summarise(q0.025 = quantile(prob, probs = 0.025),
            q0.05 = quantile(prob, probs = 0.05),
            q0.1 = quantile(prob, probs = 0.1),
            q0.25 = quantile(prob, probs = 0.25),
            q0.5 = quantile(prob, probs = 0.5),
            q0.75 = quantile(prob, probs = 0.75),
            q0.9 = quantile(prob, probs = 0.9),
            q0.95 = quantile(prob, probs = 0.95),
            q0.975 = quantile(prob, probs = 0.975))

```
**Figure 1. Rise in anti-microbial resistance.** Change in resistance over time across 13 anti-microbials. Solid lines describe the median posterior prediction, averaged across all species and countries. Error bars describe the 50%, 80%, 90% and 95% credible intervals.



```{r figure1, fig.height= 10, fig.width = 10, warning=F, message = F}
#Culumative MDR


breaks_fun <- function(x) {
  if(max(x) > 0.8){
    c(0.25, 0.5, 0.75)
  } else if (max(x) > 0.6){
    c(0.2, 0.4, 0.6)
  } else if (max(x) > 0.5){
    c(0.1, 0.3, 0.5)
  } else if (max(x) > 0.45){
    c(0.15, 0.3, 0.45)
  } else if (max(x) > 0.4){
    c(0.1, 0.25, 0.4)
  } else if (max(x) > 0.35){
    c(0.15, 0.25, 0.35)
  } else if (max(x) > 0.3){
    c(0.1, 0.2, 0.3)
  } else if (max(x) > 0.1){
    c(0.05, 0.1, 0.15)
  } else if (max(x) > 0.05){
    c(0.01, 0.03, 0.05)
  } else if (max(x) > 0.04){
    c(0.01, 0.025, 0.04)
  } else if (max(x) > 0.03){
    c(0.01, 0.02, 0.03)
  } else if (max(x) > 0.025){
    c(0.005, 0.015, 0.025)
  } else if (max(x) > 0.0125){3
    c(0.0025, 0.0075, 0.0125)
  } else if (max(x) > 0.005){
    c(0.001, 0.003, 0.005)
  } else {
    c(0.0005, 0.001, 0.0015)
  }
}

fig1 = ggplot(ab_probs) +
  geom_line(aes(x = as.numeric(year), y = q0.5), colour = "grey30") +
  geom_ribbon(aes(x = as.numeric(year), ymin = q0.25, ymax = q0.75), fill = "grey30", alpha = 0.4) +
  geom_ribbon(aes(x = as.numeric(year), ymin = q0.1, ymax = q0.9), fill = "grey30", alpha = 0.3) +
  geom_ribbon(aes(x = as.numeric(year), ymin = q0.05, ymax = q0.95), fill = "grey30", alpha = 0.2) +
  geom_ribbon(aes(x = as.numeric(year), ymin = q0.025, ymax = q0.975), fill = "grey30", alpha = 0.1) +
  scale_x_continuous(breaks = c(1990, 2005, 2020), expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0), breaks = breaks_fun) +
  facet_wrap(.~factor(ab, levels=c(
    "Oxazolidinones", 
    "Glycopeptides", 
    "Rifampicin", 
    "Quinolones", 
    "Aminoglycosides", 
    "Trimethoprim",
    "Macrolides", 
    "Fosfomycin", 
    "Sulphonamides",
    "Polymyxins", 
    "Phenicols", 
    "Tetracyclines",
    "Betalactams")), scales = "free", ncol = 3) +
  labs(x = "Year", y = "Probability of exhibiting resistance") +
  theme_classic()

fig1
```

**Figure 2. The arrival of multi-drug resistance.** Number of resistant anti-microbials (out of a maximum of 13) over time. Proportional distribution of resistance count (N) assessed over 2000 posterior samples.

```{r figure2, fig.height=8, fig.width =12, warning=F, message = F}
draw_n = 1000
tmp_df = data.frame(year = c(min(subsampledf$year): (77)))
draw = posterior_linpred(fit, tmp_df, ndraws = draw_n, re.form = NA, transform = T)
draw = matrix(draw, draw_n*111, 13)
meta = expand.grid(draws = c(1:draw_n), year = c(1990:2100))
draw = cbind(meta, draw)
colnames(draw)[3:ncol(draw)] = ab_list
draw_long = gather(draw, ab, prob, Aminoglycosides:Trimethoprim)

draw_long$bin = ifelse(draw_long$prob <= 0.5, 0, 1)

overall_mdr = draw_long %>%
  group_by(draws, year) %>%
  summarise(sum = sum(bin))

((399/326)*100)-100 #increase from 1990 to 2023 in 1 class
((126/7)*100)-100 #increase from 1990 to 2023 in 2 class

(17/1000)*100 #incidence of 3 classes
(7/1000)*100
(126/1000)*100

fig2 = ggplot(overall_mdr) +
  geom_col(aes(x = year, y = draws, fill = as.factor(sum)), width = 2, position = "fill") +
  geom_vline(aes(xintercept = 2023), linetype = "dashed") +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0), breaks = c(0,0.25, 0.5, 0.75, 1)) +
  scale_fill_manual(values = 
                         scales::seq_gradient_pal("#f5f5f5", "#9e1919", "Lab")(seq(0,1,length.out=length(unique(overall_mdr$sum)))), name = "Resistance (N)") +
  labs(x = "Year", y = "Distribution of resistance frequencies") +
  theme_classic() +
  theme(legend.position = "top") + 
  guides(fill = guide_legend(nrow = 1))
fig2
```


```{r figure3, fig.height = 11, fig.width = 9, warning=F, message = F}
tmp_df = expand.grid(year = c(-33:0), ISO.3 = unique(subsampledf$ISO.3))
tmp_df$space_id = tmp_df$ISO.3
draw = posterior_linpred(fit, tmp_df, ndraws = draw_n, 
                         re.form = ~(1 + year|ISO.3) + (1|gr(space_id, cov = space)), 
                         transform = T)
draw = matrix(draw, draw_n*nrow(tmp_df), 13)
meta = expand.grid(draw = c(1:draw_n), year = c(1990:2023), ISO.3 = unique(subsampledf$ISO.3))
draw = cbind(meta, draw)
colnames(draw)[4:ncol(draw)] = ab_list
draw_long = gather(draw, ab, prob, Aminoglycosides:Trimethoprim)
country_sum = draw_long %>%
  group_by(draw, year, ISO.3) %>%
  summarise(cumsum = sum(prob))
country_suma = subset(country_sum, year == 1990)
country_sumb = subset(country_sum, year == 2023)
country_sumb$diff = country_sumb$cumsum - country_suma$cumsum
country_sum2 = country_sumb %>%
  group_by(ISO.3) %>%
  summarise(mn_diff = mean(diff), sd_diff = sd(diff), mn_2023 = mean(cumsum), sd_2023 = sd(cumsum))

head(country_sum2[order(country_sum2$mn_2023, decreasing = TRUE), ]   )
head(country_sum2[order(country_sum2$mn_2023, decreasing = FALSE), ]   )
head(country_sum2[order(country_sum2$mn_diff, decreasing = TRUE), ]   )
head(country_sum2[order(country_sum2$mn_diff, decreasing = FALSE), ]   )


world_map = ne_countries(scale = "medium", returnclass = "sf")
merged_data = merge(world_map, country_sum2, by.x = "iso_a3", by.y = "ISO.3")
data = bi_class(merged_data, x = mn_2023, y = mn_diff, style = "quantile", dim = 4)


map = ggplot() +
  geom_sf(data = world_map, colour = "grey85", fill = NA) +
  geom_sf(data = data, mapping = aes(fill = bi_class), color = "white", size = 0.1, show.legend = FALSE) +
  bi_scale_fill(pal = palette_bi, dim = 4) +
  bi_theme() +
  theme(plot.title = element_text(size=12))

legend = bi_legend(pal = palette_bi,
                   dim = 4,
                   xlab = "Resistance (N)",
                   ylab = "Resistance \U0394\ ",
                   size = 1)

fig3a = ggdraw() +
  draw_plot(map, 0, 0, 1, 1) +
  draw_plot(legend, 0.1, 0.1, 0.35, 0.35)



#Species level MDR now and change
tmp_df = expand.grid(year = c(-33:0), species = unique(subsampledf$species))
tmp_df$phylo_id = tmp_df$species
draw = posterior_linpred(fit, tmp_df, ndraws = draw_n, 
                         re.form = ~(1 + year|species) + (1|gr(phylo_id, cov = phylo)), 
                         transform = T)
draw = matrix(draw, draw_n*nrow(tmp_df), 13)
meta = expand.grid(draw = c(1:draw_n), year = c(1990:2023), species = unique(subsampledf$species))
draw = cbind(meta, draw)
colnames(draw)[4:ncol(draw)] = ab_list
draw_long = gather(draw, ab, prob, Aminoglycosides:Trimethoprim)
species_sum = draw_long %>%
  group_by(draw, year, species) %>%
  summarise(cumsum = sum(prob))
species_suma = subset(species_sum, year == 1990)
species_sumb = subset(species_sum, year == 2023)
species_sumb$diff = species_sumb$cumsum - species_suma$cumsum
species_sum2 = species_sumb %>%
  group_by(species) %>%
  summarise(mn_diff = mean(diff), sd_diff = sd(diff), mn_2023 = mean(cumsum), sd_2023 = sd(cumsum))
trend_phy = setNames(species_sum2$mn_diff, species_sum2$species)
phytools::phylosig(tr, trend_phy, method = "lambda", test = T)

status_phy = setNames(species_sum2$mn_2023, species_sum2$species)
phytools::phylosig(tr, status_phy, method = "lambda", test = T)
species_sum2 = bi_class(species_sum2, x = mn_2023, y = mn_diff, style = "quantile", dim = 4)[,c(1,6)]

tr = tr_save
tr_trim = keep.tip(tr, unique(subsampledf$species))

colnames(species_sum2) = c("species", "A")
species_sum2$species = gsub("_", " ", species_sum2$species)
spec_simplify = str_split_fixed(species_sum2$species, " ", 2)
species_sum2$species = paste0(substr(spec_simplify[,1], 1, 1),". ", spec_simplify[,2])
species_sum2 = data.frame(species_sum2)
rownames(species_sum2) = species_sum2$species
species_sum2$species = NULL


tr = tr_save
tr_trim = keep.tip(tr, unique(df$species))
plot(tr_trim)
tr_trim$edge.length = log(tr_trim$edge.length) + abs(min(log(tr_trim$edge.length))) + 0.01
spec_simplify = str_split_fixed(tr_trim$tip.label, "_", 2)
tr_trim$tip.label = paste0(substr(spec_simplify[,1], 1, 1),". ", spec_simplify[,2])
p = ggtree(tr_trim, layout = "circular")  +
  geom_tiplab(align = T, size = 2)
p = gheatmap(p, species_sum2, offset = 50, width = 0.3, colnames = F)  +
  scale_fill_manual(values = palette_bi, name = "") +
  theme(legend.position = "none")
fig3b = open_tree(p,angle = 20)

fig3 = ggarrange(fig3a, fig3b, heights = c(0.6, 1), ncol = 1)
fig3

```

**Figure 3. Distribution of resistance over space and the tree of life.** A) Bi-scale map of resistance, describing the current frequency of resistance (in 2023; red) and change in resistance (between 1970 and 2023; blue). Frequency and change in resistance calculated as the cumulative probability of resistance across the 13 anti-microbials; B) Cumulative probability of resistance for each species across six decades: 1970, 1980, 1990, 2000, 2010, and 2020. To improve visulation, phylogentic branch lengths are presented on the log-scale.


### Load datasets

'data_gdp_wb.csv'. Metadata:
iso3 = ISO-3 country code
year = year of estimate
gdp = gdp per capita at a 2015 constant

'data_population_un.csv'. Metadata:
iso3 = ISO-3 country code
year = year of estimate
pop = total population size as of January 1st in each year
density = population density per km as July 1st in each year
pop_rate = population growth rate per year, given as a percentage
life_expectancy = life expectancy at birth (combined sexes)
birth_mortality = mortality rate (per 1000) for children under 5
migration rate = net migration per 1000 people


data_health.... Metadata:
iso3 = ISO-3 country code
year = year of estimate
health_gdp = health expenditure as a proportion of gdp
beds = hospital beds per 1,000 people
physicians = physicians per 1,000 people
undernourishment = percentage of population experiencing undernourishment
hiv = incidence of HIV per 1,000 people
malaria = incidence of malaria per 1,000 people
tuberculosis = incidence of tuberculosis per 1,000 people
antiretroviral = percetnage of people infected with HIV recieving treatment


data_environment... Metadata:
iso3 = ISO-3 country code
year = year of estimate
rural = percentage of population in rural communities
rural_growth = percentage growth in rural population
urban = percentage of population in urban communities
urban_growth = percentage growth in urban populations
air_population = PM2.5 air pollution, mean annual exposure (micrograms per cubic meter)
ag_land = Percentage of land occupied by agriculture


data_governance... Metadata:
iso3 = ISO-3 country code
year = year of estimate
gini = gini coefficient of inequality
corruption = control of corruption
gov = government effectiveness
reg = regulatory quality
acc = voice and accountability


data_anti... Metadata:
country = country name
year = year of estimate
anti = Antibiotic consumption (DDD/1,000/day)
low = lower uncertainty estimate
up = upper uncertainty estimate





```{r res_drivers, fig.height=9, fig.width =9, warning=F, message = F}
#Country level MDR now and change
base = expand.grid(year = c(1990: 2023), iso3 = unique(subsampledf$ISO.3))

coords = unique(df_save[,c("ISO.3", "centroid_lat", "centroid_lon")])
coords = coords[complete.cases(coords),]
r = getData("worldclim",var="bio",res=10)[[c(1)]]
points = SpatialPoints(data.frame(x=coords$centroid_lon,y=coords$centroid_lat), proj4string = r@crs)
coords$tavg = extract(r,points)/10


population = read.csv("data_population_un.csv")
population[population == "..."] = NA

health = read.csv("data_health_wb.csv")
health[health == ".."] = NA

enviro = read.csv("data_enviro_wb.csv")
enviro[enviro == ".."] = NA

gov = read.csv("data_governance_wb.csv")
gov[gov == ".."] = NA

gdp = read.csv("data_gdp_wb.csv")
gdp[gdp == ".."] = NA

anti = read.csv("data_anti.csv")
anti$iso3 = countrycode(anti$country, "country.name", "iso3c")
anti$country = NULL

anti2 = read.csv("data_anti2.csv")[,c(1,2,5,25)]

base = left_join(base, coords, by = c("iso3" = "ISO.3"))
base = left_join(base, population)
base = left_join(base, health)
base = left_join(base, enviro)
base = left_join(base, gov)
base = left_join(base, gdp)
base = left_join(base, anti)
base = left_join(base, anti2)


miss_row = data.frame(year = base$year, iso3 = base$iso3, na = rowSums(is.na(base)))
miss_row = miss_row %>%
  group_by(year) %>%
  summarise(na = mean(na))

figs2 = ggplot(data = miss_row) +
  geom_line(aes(x = year, y = na)) +
  labs(x = "Year", y = "Mean number of missing values") +
  theme_classic() #Majority of countires posses data in the studied timeframe


base = base %>%
  group_by(iso3) %>%
  summarise(
    lat = mean(as.numeric(centroid_lat), na.rm = T),
    lon = mean(as.numeric(centroid_lon), na.rm = T),
    density = mean(as.numeric(density), na.rm = T),
    pop_rate = mean(as.numeric(pop_rate), na.rm = T),
    life_expectancy = mean(as.numeric(life_expectancy), na.rm = T),
    migration_rate = mean(as.numeric(migration_rate), na.rm = T),
    health_gdp = mean(as.numeric(health_gdp), na.rm = T),
    beds = mean(as.numeric(beds), na.rm = T),
    physicians = mean(as.numeric(physicians), na.rm = T),
    undernourishment = mean(as.numeric(undernourishment), na.rm = T),
    antiretroviral = mean(as.numeric(antiretroviral), na.rm = T),
    hiv = mean(as.numeric(hiv), na.rm = T),
    air_pollution = mean(as.numeric(air_pollution), na.rm = T),
    ag_land = mean(as.numeric(ag_land), na.rm = T),
    gini = mean(as.numeric(gini), na.rm = T),
    corruption = mean(as.numeric(corruption), na.rm = T),
    gov = mean(as.numeric(gov), na.rm = T),
    reg = mean(as.numeric(reg), na.rm = T),
    acc = mean(as.numeric(acc), na.rm = T),
    gdp = mean(as.numeric(gdp), na.rm = T),
    anti = mean(as.numeric(anti), na.rm = T),
    tavg = mean(as.numeric(tavg), na.rm = T)
)
base = base %>% mutate_all(~ifelse(is.nan(.), NA, .))
base$gdp = log(base$gdp)
base$undernourishment = log(base$undernourishment)
base$hiv = log(base$hiv)
base$density = log(base$density)
cormat = round(cor(base[,c(4:23)], use="complete.obs"),1)
cormat = reshape2::melt(cormat)

figs3 = ggplot() + 
  geom_tile(data = cormat, aes(x=Var1, y=Var2, fill=value)) +
  scale_fill_gradient2(low = scales::muted("darkblue"), high = scales::muted("darkred"), na.value = "grey20", mid = "white", midpoint = 0, breaks = c(-1,-0.5,0,0.5,1), limits = c(-1,1), name = "Correlation\n") +
  scale_x_discrete(expand = c(0,0)) +
  scale_y_discrete(expand = c(0,0)) +
  labs(x = "", y = "") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90),
        legend.position = "bottom")

```

Using the posterior predictions of cumulative resistance for each country in the year 2023, we conducted a comparative analyses to explore assocations between resitance and a series of predictors. These predictors covered everything from antibiotic use to governance, temperature, and quality of health care. Combined, representing 23 variables. Many of these variables are highly correlated with one another which makes inference challenging. With this in mind, we explored the link between these predictors and the response variable in a simplifed linear model focussed solely on the predictors with the greatest mechansitic support: human density, propotion of GDP spent on health expeniture, governance effectiveness, antibiotic use, and the countries average temperature. We also added the following predictor interactions: health expenditure and antibiotic use, and governance effectiveness and anitbiotic use. We included these in a linear model with resistance as the response variable. We included a spatial covariance term to account for covariation in neighnouring countries.

```{r figure4, fig.height=4, fig.width =8, warning=F, message = F}

country_sum2 = left_join(country_sum2, base, by = c("ISO.3" = "iso3"))
country_sum2_t = gather(country_sum2, predictor, value, density:tavg)

figs4 = ggplot(country_sum2_t) +
  geom_point(aes(x = value, y = mn_2023)) +
  geom_smooth(aes(x = value, y = mn_2023), method = "lm") +
  labs(x = "Value", y = "Resistance frequency in 2023") +
  facet_wrap(predictor~., scales = "free") +
  theme_classic()


country_sum2 = country_sum2[complete.cases(country_sum2[,c(
  "density",
  "health_gdp",
  "gov",
  "gdp",
  "gini",
  "anti",
  "tavg")]),]

#Produce a spatial covariance matrix
spa_df = unique(country_sum2[,c("ISO.3", "lat", "lon")])
spa_mat_trim = as.matrix(distm(spa_df[,c(3,2)], fun = distHaversine))/1000000
spa_mat_trim = norm_range(spa_mat_trim)
spa_mat_trim = abs(spa_mat_trim - 1) 
spa_id = spa_df$ISO.3
colnames(spa_mat_trim) = spa_id
rownames(spa_mat_trim) = spa_id
country_sum2$space_id = country_sum2$ISO.3

plot(country_sum2$mn_2023, country_sum2$mn_diff) #Highly correlated, just use absolute mdr i.e mn_2023

form_inf = bf(mn_2023~
                scale(density) +
                scale(health_gdp) +
                scale(gov) +
                scale(anti) +
                scale(tavg) +
                scale(anti)*scale(health_gdp) +
                scale(anti)*scale(gov) +
                (1|gr(space_id, cov = space)))

m_inf = brm(
  formula = form_inf,
  data = country_sum2,
  data2 = list(space = spa_mat_trim), 
  family = gaussian,
  chains = 4, cores = 4, iter = 2000, warmup = 1000)

figs5 = pp_check(m_inf) + 
  theme_classic()
model_check = DHARMa::createDHARMa(
  simulatedResponse = (t(posterior_predict(m_inf))),
  observedResponse = country_sum2$mn_2023,
  fittedPredictedResponse = (apply(t(posterior_epred(m_inf)), 1, mean)),
  integerResponse = FALSE)
figs6 = plot(model_check)

summary(m_inf)

post_sum = as.data.frame(posterior_summary(m_inf, probs = c(0.025, 0.1, 0.25, 0.75, 0.9, 0.975))[c(2:8),])
post_sum$var = c(
  "Population density",
  "Health expenditure",
  "Governance",
  "Antibiotic-use",
  "Average temperature",
  "Health expenditure:\nAntibiotic-use",
  "Governance:\nAntibiotic-use"
)
post_sum$col = c(
  "A",
  "A",
  "B",
  "A",
  "B",
  "A",
  "A"
)

fig4a = ggplot(post_sum) +
  geom_linerange(aes(y = reorder(var, -Q2.5), xmin = Q2.5, xmax = Q97.5, colour = col), linewidth = 5, alpha = 0.4) +
  geom_linerange(aes(y = reorder(var, -Q2.5), xmin = Q10, xmax = Q90, colour = col), linewidth = 5, alpha = 0.6) +
  geom_linerange(aes(y = reorder(var, -Q2.5), xmin = Q25, xmax = Q75, colour = col), linewidth = 5, alpha = 0.8) +
  geom_vline(aes(xintercept = 0), linetype = "dashed") + 
  scale_colour_manual(guide = "none", values = c("grey30", scales::muted("#9e1919"))) +
  labs(x = expression(beta), y= "") +
  theme_classic()


pred <- predictions(m_inf,
                    newdata = datagrid(gov = seq(-2.5, 2.5, by = 0.05))) |>
  posterior_draws()

fig4b = ggplot(pred, aes(x = gov, y = exp(draw))) +
  stat_lineribbon() +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  scale_fill_manual(guide = "none",values = c(fill_alpha("#9e1919", 0.4), fill_alpha("#9e1919", 0.6), fill_alpha("#9e1919", 0.8))) +
  labs(x = "Government effectivness",
       y = expression("Resistance (N)")) +
  theme_classic()

pred <- predictions(m_inf,
                    newdata = datagrid(tavg = seq(0, 30, by = 0.1))) |>
  posterior_draws()

fig4c = ggplot(pred, aes(x = tavg, y = exp(draw))) +
  stat_lineribbon() +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  scale_fill_manual(guide = "none",values = c(fill_alpha("#9e1919", 0.4), fill_alpha("#9e1919", 0.6), fill_alpha("#9e1919", 0.8))) +
  labs(x = "Average temperature",
       y = expression("Resistance (N)")) +
  theme_classic()


ggarrange(fig4a, fig4b, fig4c, ncol = 3, widths = c(1,0.7,0.7))



```

**Figure 4. Human density, health expenditure, and financial inequality are linked to resistance ** Marginal effect posterior predictions of the association between resistance and the following predictors: human density, health expenditure, and financial inequality. Error bars describe 95% credible intervals.





### Supplementary material


```{r figure_s1, fig.height=14, fig.width =11, warning=F, message = F, echo = F}
ggarrange(plotlist = pp_list, ncol = 2, nrow = 7)
```

**Figure S1. Posterior predictive checks of temporal trend model ** 


```{r figure_s2, fig.height=4, fig.width =8, warning=F, message = F, echo = F}
figs2
```

**Figure S2. Percentage of predictors missing values predictors over time ** 


```{r figure_s3, fig.height=9, fig.width =9, warning=F, message = F, echo = F}
figs3
```

**Figure S3. Pairwise correlation in predictors ** 


```{r figure_s4, fig.height=11, fig.width =8, warning=F, message = F, echo = F}
figs4
```

**Figure S4. Univriate association between resistance and each predictor ** 


```{r figure_s5, fig.height=4, fig.width =8, warning=F, message = F, echo = F}
figs5
```

**Figure S5. Posterior predictive check ** 


```{r figure_s6, fig.height=5, fig.width =8, warning=F, message = F, echo = F}
figs6
```

**Figure S6. Assumption checking ** 

```{r robitt_compare_systematic, fig.height=5, fig.width =8, warning=F, message = F}
library(arrow)
library(brms)
library(dplyr)
library(ggrepel)

df = df_save
colnames(df)[10] = "Betalactams"

struc = read.csv("struc.csv")
struc$ind = 1
struc$id = as.character(struc$id)
#Test 1 - C. jejuni
tmp = subset(df, species == "Campylobacter jejuni")
tmp = left_join(tmp, struc)

struc_tmp = subset(tmp, ind == 1)
struc_tmp$year = struc_tmp$year - 2023 #2023 - 20 etc
range(struc_tmp$year) #2023 - 20 etc
unique(struc_tmp$ISO.3) #GBR
tmp = subset(tmp, !is.na(year))
tmp$year = tmp$year - 2023
tmp = subset(tmp, ISO.3 == "GBR")
tmp = subset(tmp, year >= -20)
tmp = subset(tmp, year <= -6)


struc_isolate = subset(tmp, ind == 1)
struc_isolate$uni_id = 1:nrow(struc_isolate)
nstruc_isolate = subset(tmp, is.na(ind))
nstruc_isolate$uni_id = 1:nrow(nstruc_isolate)

form =
  bf(mvbind(
    Aminoglycosides, 
    Betalactams, 
    Polymyxins, 
    Fosfomycin, 
    Glycopeptides, 
    Macrolides, 
    Oxazolidinones, 
    Phenicols, 
    Quinolones, 
    Rifampicin, 
    Sulphonamides,
    Tetracyclines,
    Trimethoprim) ~ 
      year 
  )  + set_rescor(FALSE)


#fit_struc = brm(form, 
#           data = struc_isolate, 
#           chains = 4, cores = 4, iter = 1000, warmup = 500, thin = 1, refresh = 10,
#           family = bernoulli)
#saveRDS(fit_struc, "fit_struc1.rds")
fit_struc = readRDS("fit_struc1.rds")

#fit_nstruc = brm(form, 
#          data = nstruc_isolate, 
#          chains = 4, cores = 4, iter = 1000, warmup = 500, thin = 1, refresh = 10,
#          family = bernoulli)
#saveRDS(fit_nstruc, "fit_nstruc1.rds")
fit_nstruc = readRDS("fit_nstruc1.rds")



nd = tmp[c(1:15),c(10:24)]
nd$year = c(-20:-6)
nd[,c(2:14)] = NA

pred1 = posterior_linpred(fit_struc, newdata = nd, transform = T, re.form = NA)
pred2 = posterior_linpred(fit_nstruc, newdata = nd, transform = T, re.form = NA)

nam = c("Aminoglycosides", "Betalactams", "Polymyxins", "Fosfomycin", "Glycopeptides", "Macrolides", "Oxazolidinones", "Phenicols", "Quinolones", "Rifampicin", "Sulphonamides","Tetracyclines","Trimethoprim")

pred_cmb = NULL
pred_cmb2 = NULL
for(a in c(1:13)){
  pred1_tmp = pred1[,,a]
  pred2_tmp = pred2[,,a]
  sum_dat = data.frame(
    cat = nam[a],
    x =  c(2003:2017),
    med1 = apply(pred1_tmp, 2, quantile, probs = 0.5),
    high1 = apply(pred1_tmp, 2, quantile, probs = 0.75),
    low1 = apply(pred1_tmp, 2, quantile, probs = 0.25),
    med2 = apply(pred2_tmp, 2, quantile, probs = 0.5),
    high2 = apply(pred2_tmp, 2, quantile, probs = 0.75),
    low2 = apply(pred2_tmp, 2, quantile, probs = 0.25))
  pred_cmb = rbind(pred_cmb, sum_dat)
  
}


ggplot(pred_cmb) +
  geom_line(aes(x = x, y = med1), alpha = 0.3) +
  geom_line(aes(x = x, y = med2), colour = "darkred", alpha = 0.3) +
  geom_ribbon(aes(x = x, ymin = low1, ymax = high1), alpha = 0.3) +
  geom_ribbon(aes(x = x, ymin = low2, ymax = high2), fill = "darkred", alpha = 0.3) +
  labs(x = "Year", y = "Probability of resistance") +
  theme_classic() +
  facet_wrap(cat~., scales = "free")

slp_df = data.frame(
  fix1 = summary(fit_struc)$fixed[c(14:26),1],
  fix2 = summary(fit_nstruc)$fixed[c(14:26),1],
  names = nam)

ggplot(data = slp_df, aes(x = fix1, y = fix2)) +
  geom_point() +
  geom_abline(aes(intercept = 0, slope = 1), linetype = "dashed") +
  labs(x = "Structured Beta", y = "Unstructured Beta") +
  theme_classic()



#Test 2 - S. pneumoniae
tmp = subset(df, species == "Streptococcus pneumoniae")
tmp = left_join(tmp, struc)
tmp$ISO.3 = ifelse(tmp$country == "Morocco", "MAR", tmp$ISO.3)#Morocco country codes are incorrect and represent month of year.

struc_tmp = subset(tmp, ind == 1)
struc_tmp$year = struc_tmp$year - 2023
range(struc_tmp$year, na.rm = T) 
unique(struc_tmp$ISO.3) 
tmp = subset(tmp, !is.na(year))
tmp = tmp[which(tmp$ISO.3 %in% unique(struc_tmp$ISO.3)),]
tmp$year = tmp$year - 2023 #2023 - 20 etc
tmp = subset(tmp, year >= -39)
tmp = subset(tmp, year <= -6)


struc_isolate = subset(tmp, ind == 1)
struc_isolate$uni_id = 1:nrow(struc_isolate)

#Produce a spatial covariance matrix
spa_df = unique(struc_isolate[,c("ISO.3", "centroid_lat", "centroid_lon")])
spa_df = subset(spa_df, !is.na(spa_df$centroid_lat))
spa_mat_trim = as.matrix(distm(spa_df[,c(3,2)], fun = distHaversine))/1000000
spa_mat_trim = norm_range(spa_mat_trim)
spa_mat_trim = abs(spa_mat_trim - 1) 
spa_id = spa_df$ISO.3
colnames(spa_mat_trim) = spa_id
rownames(spa_mat_trim) = spa_id

struc_isolate$space_id = struc_isolate$ISO.3



form =
  bf(mvbind(
    Aminoglycosides, 
    Betalactams, 
    Polymyxins, 
    Fosfomycin, 
    Glycopeptides, 
    Macrolides, 
    Oxazolidinones, 
    Phenicols, 
    Quinolones, 
    Rifampicin, 
    Sulphonamides,
    Tetracyclines,
    Trimethoprim) ~ 
      year + 
      (1 + year|ISO.3) +
      (1|gr(space_id, cov = space)) 
  )  + set_rescor(FALSE)

#fit_struc = brm(form, 
#                data = struc_isolate, 
#                data2 = list(space = spa_mat_trim), 
#                chains = 4, cores = 4, iter = 1000, warmup = 500, thin = 1, refresh = 10,
#                family = bernoulli)
#saveRDS(fit_struc, "fit_struc2.rds")
fit_struc = readRDS("fit_struc2.rds")

nstruc_isolate = subset(tmp, is.na(ind))
nstruc_isolate$uni_id = 1:nrow(nstruc_isolate)

#Produce a spatial covariance matrix
nstruc_isolate = subset(nstruc_isolate, nstruc_isolate$ISO.3 != "MAR")
spa_df = unique(nstruc_isolate[,c("ISO.3", "centroid_lat", "centroid_lon")])
spa_df = subset(spa_df, !is.na(spa_df$centroid_lat))
spa_mat_trim = as.matrix(distm(spa_df[,c(3,2)], fun = distHaversine))/1000000
spa_mat_trim = norm_range(spa_mat_trim)
spa_mat_trim = abs(spa_mat_trim - 1) 
spa_id = spa_df$ISO.3
colnames(spa_mat_trim) = spa_id
rownames(spa_mat_trim) = spa_id

nstruc_isolate$space_id = nstruc_isolate$ISO.3

#fit_nstruc = brm(form, 
#                 data = nstruc_isolate, 
#                 data2 = list(space = spa_mat_trim), 
#                 chains = 4, cores = 4, iter = 1000, warmup = 500, thin = 1, refresh = 10,
#                 family = bernoulli)
#saveRDS(fit_nstruc, "fit_nstruc2.rds")
fit_nstruc = readRDS("fit_nstruc2.rds")


nd = tmp[c(1:23),c(10:24)]
nd$year = c(-28:-6)
nd[,c(2:14)] = NA

pred1 = posterior_linpred(fit_struc, newdata = nd, transform = T, re.form = NA)
pred2 = posterior_linpred(fit_nstruc, newdata = nd, transform = T, re.form = NA)

nam = c("Aminoglycosides", "Betalactams", "Polymyxins", "Fosfomycin", "Glycopeptides", "Macrolides", "Oxazolidinones", "Phenicols", "Quinolones", "Rifampicin", "Sulphonamides","Tetracyclines","Trimethoprim")

pred_cmb = NULL
pred_cmb2 = NULL
for(a in c(1:13)){
  pred1_tmp = pred1[,,a]
  pred2_tmp = pred2[,,a]
  sum_dat = data.frame(
    cat = nam[a],
    x =  c(1995:2017),
    med1 = apply(pred1_tmp, 2, quantile, probs = 0.5),
    high1 = apply(pred1_tmp, 2, quantile, probs = 0.75),
    low1 = apply(pred1_tmp, 2, quantile, probs = 0.25),
    med2 = apply(pred2_tmp, 2, quantile, probs = 0.5),
    high2 = apply(pred2_tmp, 2, quantile, probs = 0.75),
    low2 = apply(pred2_tmp, 2, quantile, probs = 0.25))
  pred_cmb = rbind(pred_cmb, sum_dat)

}

ggplot(pred_cmb) +
  geom_line(aes(x = x, y = med1), alpha = 0.3) +
  geom_line(aes(x = x, y = med2), colour = "darkred", alpha = 0.3) +
  geom_ribbon(aes(x = x, ymin = low1, ymax = high1), alpha = 0.3) +
  geom_ribbon(aes(x = x, ymin = low2, ymax = high2), fill = "darkred", alpha = 0.3) +
  labs(x = "Year", y = "Probability of resistance") +
  theme_classic() +
  facet_wrap(cat~., scales = "free")

slp_df = data.frame(
  fix1 = summary(fit_struc)$fixed[c(14:26),1],
  fix2 = summary(fit_nstruc)$fixed[c(14:26),1],
  names = nam)

ggplot(data = slp_df, aes(x = fix1, y = fix2)) +
  geom_point() +
  geom_abline(aes(intercept = 0, slope = 1), linetype = "dashed") +
  labs(x = "Structured Beta", y = "Unstructured Beta") +
  theme_classic()




#Test 3 - Neisseria gonorrhoeae
tmp = subset(df, species == "Neisseria gonorrhoeae")
tmp = left_join(tmp, struc)
tmp$year = tmp$year - 2023

struc_tmp = subset(tmp, ind == 1)
range(struc_tmp$year, na.rm = T) #2023 - 20 etc
tmp = subset(tmp, !is.na(year))
unique(struc_tmp$ISO.3) 
struc_tmp = subset(struc_tmp, struc_tmp$ISO.3 != "UGANDA")
struc_tmp = subset(struc_tmp, struc_tmp$ISO.3 != "")
struc_tmp = subset(struc_tmp, struc_tmp$ISO.3 != "JUL")
struc_tmp = subset(struc_tmp, struc_tmp$ISO.3 != "SEP")
tmp = tmp[which(tmp$ISO.3 %in% unique(struc_tmp$ISO.3)),]
tmp = subset(tmp, year >= -33) #min 1990
tmp = subset(tmp, year <= -5)


struc_isolate = subset(tmp, ind == 1)
struc_isolate$uni_id = 1:nrow(struc_isolate)

#Produce a spatial covariance matrix
spa_df = unique(struc_isolate[,c("ISO.3", "centroid_lat", "centroid_lon")])
spa_mat_trim = as.matrix(distm(spa_df[,c(3,2)], fun = distHaversine))/1000000
spa_mat_trim = norm_range(spa_mat_trim)
spa_mat_trim = abs(spa_mat_trim - 1) 
spa_id = spa_df$ISO.3
colnames(spa_mat_trim) = spa_id
rownames(spa_mat_trim) = spa_id

struc_isolate$space_id = struc_isolate$ISO.3



form =
  bf(mvbind(
    Aminoglycosides, 
    Betalactams, 
    Polymyxins, 
    Fosfomycin, 
    Glycopeptides, 
    Macrolides, 
    Oxazolidinones, 
    Phenicols, 
    Quinolones, 
    Rifampicin, 
    Sulphonamides,
    Tetracyclines,
    Trimethoprim) ~ 
      year + 
      (1 + year|ISO.3) +
      (1|gr(space_id, cov = space)) 
  )  + set_rescor(FALSE)

#fit_struc = brm(form, 
#                data = struc_isolate, 
#                data2 = list(space = spa_mat_trim), 
#                chains = 4, cores = 4, iter = 1000, warmup = 500, thin = 1, refresh = 10,
#                family = bernoulli)
#saveRDS(fit_struc, "fit_struc3.rds")
fit_struc = readRDS("fit_struc3.rds")

nstruc_isolate = subset(tmp, is.na(ind))
nstruc_isolate$uni_id = 1:nrow(nstruc_isolate)

#Produce a spatial covariance matrix
spa_df = unique(nstruc_isolate[,c("ISO.3", "centroid_lat", "centroid_lon")])
spa_mat_trim = as.matrix(distm(spa_df[,c(3,2)], fun = distHaversine))/1000000
spa_mat_trim = norm_range(spa_mat_trim)
spa_mat_trim = abs(spa_mat_trim - 1) 
spa_id = spa_df$ISO.3
colnames(spa_mat_trim) = spa_id
rownames(spa_mat_trim) = spa_id

nstruc_isolate$space_id = nstruc_isolate$ISO.3

#fit_nstruc = brm(form, 
#                 data = nstruc_isolate, 
#                 data2 = list(space = spa_mat_trim), 
#                 chains = 4, cores = 4, iter = 1000, warmup = 500, thin = 1, refresh = 10,
#                 family = bernoulli)
#saveRDS(fit_nstruc, "fit_nstruc3.rds")
fit_nstruc = readRDS("fit_nstruc3.rds")

nd = tmp[c(1:23),c(10:24)]
nd$year = c(-28:-6)
nd[,c(2:14)] = NA

pred1 = posterior_linpred(fit_struc, newdata = nd, transform = T, re.form = NA)
pred2 = posterior_linpred(fit_nstruc, newdata = nd, transform = T, re.form = NA)

nam = c("Aminoglycosides", "Betalactams", "Polymyxins", "Fosfomycin", "Glycopeptides", "Macrolides", "Oxazolidinones", "Phenicols", "Quinolones", "Rifampicin", "Sulphonamides","Tetracyclines","Trimethoprim")

pred_cmb = NULL
pred_cmb2 = NULL
for(a in c(1:13)){
  pred1_tmp = pred1[,,a]
  pred2_tmp = pred2[,,a]
  sum_dat = data.frame(
    cat = nam[a],
    x =  c(1995:2017),
    med1 = apply(pred1_tmp, 2, quantile, probs = 0.5),
    high1 = apply(pred1_tmp, 2, quantile, probs = 0.75),
    low1 = apply(pred1_tmp, 2, quantile, probs = 0.25),
    med2 = apply(pred2_tmp, 2, quantile, probs = 0.5),
    high2 = apply(pred2_tmp, 2, quantile, probs = 0.75),
    low2 = apply(pred2_tmp, 2, quantile, probs = 0.25))
  pred_cmb = rbind(pred_cmb, sum_dat)

}

ggplot(pred_cmb) +
  geom_line(aes(x = x, y = med1), alpha = 0.3) +
  geom_line(aes(x = x, y = med2), colour = "darkred", alpha = 0.3) +
  geom_ribbon(aes(x = x, ymin = low1, ymax = high1), alpha = 0.3) +
  geom_ribbon(aes(x = x, ymin = low2, ymax = high2), fill = "darkred", alpha = 0.3) +
    labs(x = "Year", y = "Probability of resistance") +
  theme_classic() +
  facet_wrap(cat~., scales = "free")

slp_df = data.frame(
  fix1 = summary(fit_struc)$fixed[c(14:26),1],
  fix2 = summary(fit_nstruc)$fixed[c(14:26),1],
  names = nam)

ggplot(data = slp_df, aes(x = fix1, y = fix2)) +
  geom_point() +
  geom_abline(aes(intercept = 0, slope = 1), linetype = "dashed") +
  labs(x = "Structured Beta", y = "Unstructured Beta") +
  theme_classic()

```